---
title: "M05-Visualizing Data with Tables"
author: "Daisy Hernandez"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    toc-depth: 3
    number-sections: true
    code-link: true
    code-fold: show
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute: 
  freeze: auto
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Reflection Essay {#sec-essay}

::: callout-note
**Prompt:** Summarize what you learned from the videos you watched. What did you like about the `gt` and `gtExtra` packages demonstrated in the videos? How do tables complement charts for data visualization? Under what circumstances would you prefer to use tables rather than charts to visualize data?
:::

This module's videos walked through how the `gt` package turns a plain data frame into a polished, presentation-ready table by layering on titles, spanner labels, footnotes, and source notes, while `gtExtra` builds on top of that with pre-built themes and inline mini-visuals like sparklines and bar charts embedded directly in table cells. What stood out most was how much control `gt` gives over labeling and formatting without writing custom HTML or CSS by hand.

I liked that `gtExtra` in particular makes it fast to add things like inline bar charts (`gt_plt_bar`) or sparklines (`gt_plt_sparkline`) directly into a cell, which lets a single table communicate both an exact value and a trend at a glance. That combination is something a static chart alone cannot do as compactly.

> Tables and charts complement each other because charts are best at showing overall patterns, trends, and relative comparisons at a glance, while tables are best at communicating precise, individual values that a reader may need to look up, audit, or cite directly.

I would prefer a table over a chart when the audience needs exact numbers (e.g., dollar figures for a budget review), when there are many categories that would clutter a chart legend, or when the report needs to double as a reference document rather than just a visual takeaway.

# Building a gt/gtExtra Table {#sec-table}

## Load Packages and Data

```{r libraries}
library(tidyverse)
library(gt)
library(gtExtras)
```

```{r data}
# Synthetic direct-mail marketing dataset, from this week's lecture
# (jaejungca.github.io/recipes, W02 — Data Splitting and Preprocessing)
set.seed(123)
n <- 3000

customer_data <- tibble(
  customer_id = paste0("C", str_pad(1:n, 5, pad = "0")),
  age = round(rnorm(n, mean = 45, sd = 12)),
  income = round(rlnorm(n, meanlog = 10.8, sdlog = 0.6)), # right-skewed
  recency_days = round(rexp(n, rate = 1 / 60)), # days since last purchase
  freq_12mo = rpois(n, lambda = 3), # purchases in 12 months
  avg_order_amt = round(rlnorm(n, meanlog = 4.2, sdlog = 0.5), 2),
  channel = sample(
    c("email", "direct_mail", "digital"),
    n, replace = TRUE, prob = c(0.5, 0.3, 0.2)
  ),
  region = sample(c("West", "South", "Midwest", "Northeast"), n, replace = TRUE),
  loyalty_tier = sample(
    c("Bronze", "Silver", "Gold", "Platinum"),
    n, replace = TRUE, prob = c(0.4, 0.3, 0.2, 0.1)
  ),
  responded = rbinom(
    n, 1,
    prob = plogis(
      -3 +
        0.02 * (age - 45) +
        0.3 * log(income / 50000) +
        0.1 * freq_12mo -
        0.005 * recency_days
    )
  )
) |>
  mutate(
    income = if_else(runif(n) < 0.08, NA_real_, income),
    avg_order_amt = if_else(runif(n) < 0.05, NA_real_, avg_order_amt),
    responded = factor(responded, levels = c(1, 0), labels = c("yes", "no"))
  )
```

## Wrangle Data

```{r wrangle}
summary_table <- customer_data %>%
  group_by(region, loyalty_tier) %>%
  summarise(
    avg_income = mean(income, na.rm = TRUE),
    avg_order = mean(avg_order_amt, na.rm = TRUE),
    avg_freq = mean(freq_12mo),
    response_rate = mean(responded == "yes"),
    n_customers = n(),
    .groups = "drop"
  ) %>%
  arrange(region, loyalty_tier)
```

## Build the Table

```{r gt-table}
summary_table %>%
  gt(groupname_col = "region", rowname_col = "loyalty_tier") %>%
  gt_theme_espn() %>%
  tab_header(
    title = "Customer Profile and Response Rate by Region and Loyalty Tier",
    subtitle = "Synthetic direct-mail dataset — M05 Table Visualization"
  ) %>%
  tab_stubhead(label = "Region / Loyalty Tier") %>%
  cols_label(
    avg_income = "Avg. Income",
    avg_order = "Avg. Order Amount",
    avg_freq = "Avg. Purchases (12mo)",
    response_rate = "Response Rate",
    n_customers = "# Customers"
  ) %>%
  tab_spanner(
    label = "Purchase Behavior",
    columns = c(avg_order, avg_freq)
  ) %>%
  fmt_currency(columns = c(avg_income, avg_order), currency = "USD", decimals = 2) %>%
  fmt_number(columns = avg_freq, decimals = 2) %>%
  fmt_percent(columns = response_rate, decimals = 1) %>%
  data_color(
    columns = response_rate,
    colors = scales::col_numeric(
      palette = c("#FDE0DD", "#C51B8A"),
      domain = NULL
    )
  ) %>%
  tab_source_note(source_note = "Source: Synthetic direct-mail dataset, IBM 6540 W02 lecture materials (jaejungca.github.io/recipes).") %>%
  tab_footnote(
    footnote = "Response rate = share of customers in that segment who responded 'yes' to the direct-mail campaign.",
    locations = cells_column_labels(columns = response_rate)
  ) %>%
  tab_options(
    table.font.size = px(13),
    heading.title.font.size = px(20)
  )
```

# Why This Table Matters for the MSDM Project {#sec-relevance}

::: callout-note
**Prompt:** Explain why the table is worth adding to your MSDM project written report or presentation slides.
:::

A table like this earns a place in the MSDM project report because it lets a stakeholder audit exact figures (spend, visit frequency, satisfaction) by segment rather than eyeballing bar heights, which matters when the same numbers may inform a budget or targeting decision. It also compresses several metrics and segments into one reference-friendly object instead of requiring three or four separate charts, which keeps the report tighter without losing the underlying detail.

# Published GitHub Page {#sec-publish}

*Add your published GitHub Pages URL here after publishing the rendered HTML.*